---
title: "ST606 Final Project — Bank Term Deposit Subscription Analysis"
author: "Abhimanyu Muraleeharan Sreekala (25254325)"
date: "2026-03-12"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

# Introduction

This project analyses the Bank Marketing dataset to determine which client and
campaign features are associated with subscription to a term deposit. A secondary
research question investigates whether call duration is significantly associated
with the probability of subscription. Binary logistic regression is the primary
modelling technique employed.

---

# Data Loading & Setup

## Load Libraries

```{r}
#install.packages("pROC")
#install.packages("caret")
#install.packages("car")
#install.packages("arm")
#install.packages("ResourceSelection")
##install.packages("vivid")
#install.packages("ranger")
#install.packages("glmnet")
#install.packages("rlang")
#install.packages("factoextra")
#install.packages("reshape2")
#install.packages("smotefamily")
#install.packages(c("patchwork", "cowplot"))
```


```{r libraries}
library(tidyverse)
library(ggplot2)
library(reshape2)
library(MASS)
library(pROC)
library(caret)
library(car)
library(arm)
library(ResourceSelection)
library(vivid)
library(ranger)
library(dplyr)
library(patchwork)
library(cowplot)
```

## Load Data

```{r load-data}
bank <- read.csv("/Users/abhi/Rstudio/ST606_Final_Project/bank-full.csv", sep = ";")
```

---

# Exploratory Data Analysis

## Initial Inspection

```{r inspect-structure}
str(bank)
```

```{r inspect-dims}
dim(bank)
```

```{r inspect-names}
names(bank)
```

```{r inspect-head}
head(bank)
```

```{r inspect-tail}
tail(bank)
```

```{r inspect-summary}
summary(bank)
```

```{r missing-values}
colSums(is.na(bank))
```

> There are **no missing values** in this dataset — all columns return 0.

## Outcome Variable Distribution

```{r outcome-table}
table(bank$y)
prop.table(table(bank$y))
```

```{r outcome-plot}
ggplot(bank, aes(x = y, fill = y)) +
  geom_bar() +
  geom_text(stat = "count",
            aes(label = paste0(round(after_stat(count) / nrow(bank) * 100, 1), "%")),
            vjust = -0.5, size = 4) +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  labs(title = "Distribution of Term Deposit Subscription",
       x = "Subscribed", y = "Count") +
  theme_minimal() +
  theme(legend.position = "none")
```

> The outcome variable `y` is **heavily imbalanced**: approximately 88.3% of clients
> did not subscribe (`no`) and only 11.7% subscribed (`yes`). This class imbalance
> must be addressed during modelling.

## Data Type Conversion

```{r convert-factors}
bank <- bank %>%
  mutate(across(where(is.character), as.factor))

sapply(bank, class)
sapply(bank, nlevels)
```

## Variable Descriptions

```{r variable-description}
variable_description <- data.frame(
  Variable = names(bank),
  Type = sapply(bank, class),
  Description = c(
    "Age of client",
    "Type of job",
    "Marital status",
    "Education level",
    "Credit in default",
    "Average yearly balance (EUR)",
    "Housing loan",
    "Personal loan",
    "Contact communication type",
    "Last contact day",
    "Last contact month",
    "Call duration in seconds",
    "Number of campaign contacts",
    "Days since previous contact",
    "Number of previous contacts",
    "Previous campaign outcome",
    "Subscribed term deposit (outcome)"
  )
)

variable_description
```

---

# Descriptive Statistics

## Numeric Variables

```{r numeric-summary}
numeric_vars <- bank[sapply(bank, is.numeric)]

descriptive_table <- data.frame(
  Mean   = sapply(numeric_vars, mean),
  SD     = sapply(numeric_vars, sd),
  Min    = sapply(numeric_vars, min),
  Median = sapply(numeric_vars, median),
  Max    = sapply(numeric_vars, max)
)

round(descriptive_table, 2)
```

## Categorical Variables — Frequency Tables

```{r freq-tables}
round(prop.table(table(bank$job))      * 100, 2)
round(prop.table(table(bank$education))* 100, 2)
round(prop.table(table(bank$contact))  * 100, 2)
round(prop.table(table(bank$poutcome)) * 100, 2)
round(prop.table(table(bank$y))        * 100, 2)
```

> - **Job:** The most common occupations are blue-collar, management, technician, and admin. The least common are unknown, student, housemaid, and entrepreneur.
> - **Education:** Secondary education is most prevalent, followed by tertiary and then primary".
> - **Contact:** Cellular contact is most common, accounting for the majority of contacts.
> - **Previous outcome (`poutcome`):** The majority are "unknown", followed by "failure" as the most prominent known outcome.
> - **Subscription (`y`):** "No" is the dominant outcome (~88%), reflecting a significant class imbalance.

---

# Univariate Visualisation

## Histograms — Continuous Variables

```{r hist-age}
hist(bank$age,
     main = "Histogram of Age",
     xlab = "Age",
     breaks = 20)
```

> Roughly right-skewed with a slight bell shape; highest frequency between ages 30–35.

```{r hist-balance}
hist(bank$balance,
     main = "Histogram of Balance",
     xlab = "Balance (EUR)",
     breaks = 30)
```

> Strongly right-skewed, with most clients holding lower balances and a long upper tail of high-balance outliers.

```{r hist-duration}
hist(bank$duration,
     main = "Histogram of Call Duration",
     xlab = "Duration (seconds)",
     breaks = 30)
```

> Strongly right-skewed — most calls are short, but a few very long calls pull the tail upward.

```{r hist-campaign}
hist(bank$campaign,
     main = "Histogram of Campaign Contacts",
     xlab = "Number of Contacts",
     breaks = 20)
```

> Heavily right-skewed; most clients were contacted only 1–3 times during the campaign.

```{r hist-previous}
hist(bank$previous,
     main = "Histogram of Previous Contacts",
     xlab = "Previous",
     breaks = 20)
```

> Very strongly right-skewed — the vast majority of clients had zero previous contacts before this campaign.

## Boxplots — Continuous Variables

```{r box-age}
boxplot(bank$age,
        main = "Boxplot of Age",
        ylab = "Age")
```

> The median age is approximately 39. Many outliers appear above the upper whisker (ages up to ~95), confirming a positively skewed distribution.

```{r box-balance}
boxplot(bank$balance,
        main = "Boxplot of Balance",
        ylab = "Balance (EUR)")
```

> Median is near the lower portion of the box, with numerous high-value outliers reaching ~100,000 EUR. Highly positively skewed with a few negative-balance clients as well.

```{r box-duration}
boxplot(bank$duration,
        main = "Boxplot of Call Duration",
        ylab = "Duration (seconds)")
```

> Many upper outliers including extreme values approaching 5,000 seconds. Strongly positively skewed with no notable lower outliers.

```{r box-campaign}
boxplot(bank$campaign,
        main = "Boxplot of Campaign Contacts",
        ylab = "Number of Contacts")
```

> The median is approximately 2, with many upper outliers up to ~63 contacts. Positively skewed distribution.

```{r box-previous}
boxplot(bank$previous,
        main = "Boxplot of Previous Contacts",
        ylab = "Previous")
```

> The median is very close to 0. There are numerous upper outliers, including one extreme value around 275. Highly positively skewed.

---

# Bivariate Analysis

## Correlation Matrix — Numeric Variables

```{r correlation-matrix}
cor_matrix <- cor(numeric_vars)
round(cor_matrix, 2)
```

```{r correlation-heatmap}
cor_melt <- melt(cor_matrix)

ggplot(cor_melt, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "tomato",
                       midpoint = 0, limits = c(-1, 1)) +
  geom_text(aes(label = round(value, 2)), size = 3) +
  labs(title = "Correlation Heatmap — Numeric Variables",
       x = "", y = "", fill = "Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

> The strongest positive correlation is between `pdays` and `previous` (r = 0.45), indicating a moderate positive relationship — clients contacted more previously also tend to have higher pdays values. Most other correlations are close to zero, suggesting weak or negligible linear relationships among the remaining numeric predictors.

## Scatter Plots — Key Relationships

```{r scatter-pdays-previous}
plot(bank$pdays, bank$previous,
     main = "Scatter Plot: pdays vs previous",
     xlab = "pdays", ylab = "previous")
```

> A slight upward trend is consistent with the moderate positive correlation (r = 0.45). Most points cluster at low previous values, with wide spread in pdays. A few strong outliers are present.

```{r scatter-age-balance}
plot(bank$age, bank$balance,
     main = "Scatter Plot: age vs balance",
     xlab = "Age", ylab = "Balance")
```

> Widely dispersed with no clear trend, matching the very weak positive correlation (r = 0.10). Most points cluster at lower balances across all ages, with several high-balance outliers.

```{r scatter-duration-campaign}
plot(bank$duration, bank$campaign,
     main = "Scatter Plot: duration vs campaign",
     xlab = "Duration (seconds)", ylab = "Campaign")
```

> A slight downward trend, consistent with the weak negative correlation (r = -0.08). Points are heavily clustered at lower duration values.

## Categorical Variables vs Subscription (`y`)

```{r cat-vs-y}
prop.table(table(bank$job,       bank$y), margin = 1)
prop.table(table(bank$education, bank$y), margin = 1)
prop.table(table(bank$contact,   bank$y), margin = 1)
prop.table(table(bank$poutcome,  bank$y), margin = 1)
prop.table(table(bank$marital,   bank$y), margin = 1)
```

> - **Job:** Students (28.7%) and retired clients (22.8%) have the highest subscription rates; blue-collar workers (7.3%) have the lowest.
> - **Education:** Subscription rates increase with education level; tertiary-educated clients subscribe most (15.0%) and primary-educated least (8.6%).
> - **Contact:** Cellular contact yields the highest subscription rate (14.9%); unknown contact type yields the lowest (4.1%).
> - **Previous outcome:** Previous success dramatically increases subscription likelihood (64.7%), making `poutcome` one of the strongest predictors.
> - **Marital status:** Single clients subscribe most (15.0%), married clients least (10.1%); differences are moderate.

```{r subscription-rate-plots}
plot_sub_rate <- function(var, label) {
  bank %>%
    group_by(.data[[var]]) %>%
    summarise(sub_rate = mean(y == "yes") * 100, .groups = "drop") %>%
    ggplot(aes(x = reorder(.data[[var]], sub_rate), y = sub_rate)) +
    geom_col(fill = "steelblue") +
    geom_text(aes(label = paste0(round(sub_rate, 1), "%")),
              hjust = -0.1, size = 3.5) +
    coord_flip() +
    ylim(0, 70) +
    labs(title = paste("Subscription Rate by", label),
         x = label, y = "Subscription Rate (%)") +
    theme_minimal()
}

plot_sub_rate("job",       "Job Type")
plot_sub_rate("education", "Education Level")
plot_sub_rate("poutcome",  "Previous Campaign Outcome")
plot_sub_rate("marital",   "Marital Status")
plot_sub_rate("contact",   "Contact Type")
```

## Numeric Variables vs Subscription (`y`)

```{r num-vs-y}
ggplot(bank, aes(x = y, y = duration, fill = y)) +
  geom_boxplot(outlier.alpha = 0.2) +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  labs(title = "Call Duration by Subscription Outcome",
       x = "Subscribed", y = "Duration (seconds)") +
  theme_minimal() + theme(legend.position = "none")

ggplot(bank, aes(x = y, y = balance, fill = y)) +
  geom_boxplot(outlier.alpha = 0.2) +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  labs(title = "Account Balance by Subscription Outcome",
       x = "Subscribed", y = "Balance (EUR)") +
  theme_minimal() + theme(legend.position = "none")

ggplot(bank, aes(x = y, y = age, fill = y)) +
  geom_boxplot(outlier.alpha = 0.2) +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  labs(title = "Age by Subscription Outcome",
       x = "Subscribed", y = "Age") +
  theme_minimal() + theme(legend.position = "none")
```

> `duration` shows the clearest separation between outcomes — the "yes" group has a substantially higher median call duration, strongly suggesting it is a key predictor. `balance` and `age` show much smaller differences and heavy overlap between groups.

---

# Data Preparation for Modelling

## Train / Test Split

```{r train-test-split}
set.seed(123)

train_index <- sample(
  1:nrow(bank),
  size = 0.6 * nrow(bank)
)

train_data <- bank[train_index, ]

remaining <- bank[-train_index, ]

valid_index <- sample(
  1:nrow(remaining),
  size = 0.5 * nrow(remaining)
)

valid_data <- remaining[valid_index, ]
test_data  <- remaining[-valid_index, ]

dim(valid_data)
dim(test_data)
dim(train_data)
```

## Class Balance Verification: Training vs Test Data

Before modelling, it is important to confirm that the 70/30 random split preserved
the original class proportions in both subsets.

```{r class-balance}
cat("=== Full Dataset ===\n")
round(prop.table(table(bank$y)) * 100, 2)

cat("\n=== Training Data (60%) ===\n")
round(prop.table(table(train_data$y)) * 100, 2)

cat("\n=== Validating Data (20%) ===\n")
round(prop.table(table(valid_data$y)) * 100, 2)

cat("\n=== Test Data (20%) ===\n")
round(prop.table(table(test_data$y)) * 100, 2)
```
```{r}
balance_df <- data.frame(
  Split = rep(c("Full Dataset", "Training (60%)", "Validation (20%)", "Test (20%)"), each = 2),
  Outcome = rep(c("no", "yes"), times = 4),
  Pct = c(
    round(prop.table(table(bank$y))       * 100, 1)[c("no","yes")],
    round(prop.table(table(train_data$y)) * 100, 1)[c("no","yes")],
    round(prop.table(table(valid_data$y)) * 100, 1)[c("no","yes")],
    round(prop.table(table(test_data$y))  * 100, 1)[c("no","yes")]
  )
)
```

```{r class-balance-plot}
ggplot(balance_df, aes(x = Split, y = Pct, fill = Outcome)) +
  geom_col(position = "dodge") +
  geom_text(aes(label = paste0(Pct, "%")),
            position = position_dodge(width = 0.9),
            vjust = -0.4,
            size = 4.5) +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  labs(
    title = "Class Distribution Across Full Dataset, Training, Validation & Test Sets",
    x = NULL,
    y = "Percentage (%)"
  ) +
  theme_minimal() +
  ylim(0, 100) +
  theme(
    plot.title   = element_text(size = 14, face = "bold"),
    axis.text.x  = element_text(size = 13),
    axis.text.y  = element_text(size = 13),
    axis.title.y = element_text(size = 14, face = "bold"),
    legend.title = element_text(size = 14),
    legend.text  = element_text(size = 13)
  )
```

> The proportions of "yes" (~11.7%) and "no" (~88.3%) are consistent across all three
> splits. This confirms that the random split did not introduce any accidental bias in
> class representation — both the training and test sets are representative samples of
> the original dataset. This is important because the model is trained and evaluated
> under the same class imbalance conditions.

---

# Modelling

## Simple Logistic Regression — Duration Only

This model addresses the primary research question: is call duration significantly
associated with the probability of subscription?

```{r simple-model}
model_duration <- glm(y ~ duration,
                      data   = train_data,
                      family = binomial)

summary(model_duration)
```

```{r simple-model-equation}
coef_simple <- coef(model_duration)

cat("=== Simple Model Equation ===\n")
cat("log( p / (1-p) ) =", round(coef_simple[1], 4),
    "+", round(coef_simple[2], 6), "* duration\n\n")
cat("Where p = probability of subscribing to a term deposit\n")
cat("- Intercept:", round(coef_simple[1], 4), "= log-odds when duration = 0\n")
cat("- duration: each additional second increases log-odds by",
    round(coef_simple[2], 6), "\n")
```

```{r simple-model-or}
or_simple <- exp(cbind(OR = coef(model_duration), confint(model_duration)))
cat("=== Odds Ratios and 95% CI — Simple Model ===\n")
round(or_simple, 4)
```

```{r simple-model-effect}
cat("Effect of 1 extra minute of call duration:\n")
cat("OR =", round(exp(60 * coef_simple[2]), 4),
    "— odds increase by",
    round((exp(60 * coef_simple[2]) - 1) * 100, 1), "%\n")

cat("Effect of 5 extra minutes of call duration:\n")
cat("OR =", round(exp(300 * coef_simple[2]), 4),
    "— odds increase by",
    round((exp(300 * coef_simple[2]) - 1) * 100, 1), "%\n")
```

> The coefficient for `duration` is positive and highly significant (p < 2×10⁻¹⁶).
> Each additional second of call duration increases the odds of subscription by
> approximately **0.36%**. An extra minute of call duration increases the odds by
> approximately **23.7%** in the simple model. This confirms that call duration is
> significantly associated with term deposit subscription.

```{r simple-model-curve}
duration_seq <- data.frame(duration = seq(0, 1500, by = 10))
duration_seq$prob <- predict(model_duration, newdata = duration_seq, type = "response")

ggplot(duration_seq, aes(x = duration, y = prob)) +
  geom_line(color = "steelblue", linewidth = 1.2) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "tomato") +
  labs(title = "Predicted Probability of Subscription vs Call Duration",
       x = "Call Duration (seconds)",
       y = "Predicted Probability of Subscribing") +
  theme_minimal()
```

```{r simple-model-accuracy}
pred_simple_prob  <- predict(model_duration, newdata = test_data, type = "response")
pred_simple_class <- factor(ifelse(pred_simple_prob >= 0.5, "yes", "no"),
                            levels = c("no", "yes"))
test_data$y       <- factor(test_data$y, levels = c("no", "yes"))

acc_simple <- mean(pred_simple_class == test_data$y)
cat("Simple model accuracy:", round(acc_simple * 100, 2), "%\n")
```

---

## Multiple Logistic Regression — All Key Predictors

```{r multi-model}
model_multi <- glm(y ~ duration + age + balance + campaign + previous +
                     job + marital + education + contact + poutcome,
                   data   = train_data,
                   family = binomial)

summary(model_multi)
```

```{r multi-model-or}
or_multi    <- exp(cbind(OR = coef(model_multi), confint(model_multi)))
or_multi_df <- as.data.frame(round(or_multi, 4))
or_multi_df$Predictor  <- rownames(or_multi_df)
or_multi_df            <- or_multi_df[, c("Predictor", "OR", "2.5 %", "97.5 %")]
colnames(or_multi_df)  <- c("Predictor", "OR", "CI_lower", "CI_upper")
or_multi_df$Significant <- ifelse(or_multi_df$CI_lower > 1 |
                                    or_multi_df$CI_upper < 1, "Yes", "No")
print(or_multi_df)
```
```{r}
multi_or <- exp(cbind(
  OR = coef(model_multi),
  confint(model_multi)
))

multi_table <- data.frame(
  Predictor = rownames(multi_or),
  OR        = round(multi_or[,1], 3),
  CI        = paste0(
                "(",
                round(multi_or[,2], 3),
                ", ",
                round(multi_or[,3], 3),
                ")"
              ),
  p_value   = round(summary(model_multi)$coefficients[,4], 4)
)

multi_table <- multi_table[-1, ]   # remove intercept

print(multi_table)
```
```{r}
coef_duration <- coef(model_multi)["duration"]

ci_duration <- confint(model_multi)["duration", ]

OR_1min <- exp(60 * coef_duration)

CI_1min <- exp(60 * ci_duration)

OR_1min
CI_1min
```

```{r multi-model-comparison}
cat("=== Model Comparison: Simple vs Multiple ===\n")
cat("Simple model  AIC:", round(AIC(model_duration), 2), "\n")
cat("Multiple model AIC:", round(AIC(model_multi), 2), "\n\n")

anova(model_duration, model_multi, test = "Chisq")

mcfadden_simple <- 1 - (model_duration$deviance / model_duration$null.deviance)
mcfadden_multi  <- 1 - (model_multi$deviance    / model_multi$null.deviance)

cat("McFadden R² — Simple model: ",  round(mcfadden_simple, 4), "\n")
cat("McFadden R² — Multiple model:", round(mcfadden_multi,  4), "\n")
```

> Adding predictors substantially improves model fit. The multiple model achieves a
> lower AIC and a higher McFadden R² (0.2925 vs 0.1588), and the likelihood ratio test
> confirms the improvement is statistically significant.

```{r multi-model-or-plot}
or_plot_df <- or_multi_df[-1, ]  # remove intercept

ggplot(or_plot_df, aes(x = reorder(Predictor, OR),
                        y = OR, color = Significant)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.3) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey40") +
  coord_flip() +
  scale_color_manual(values = c("Yes" = "steelblue", "No" = "tomato")) +
  labs(title = "Odds Ratios with 95% CI — Multiple Model",
       x = "Predictor", y = "Odds Ratio", color = "Significant") +
  theme_minimal()
```

---

## Variable Selection — Stepwise AIC

```{r stepwise}
model_step <- stepAIC(model_multi, direction = "both", trace = FALSE)
summary(model_step)
```

```{r stepwise-comparison}
cat("=== Variables in Multiple Model ===\n")
print(names(coef(model_multi)))

cat("\n=== Variables Retained After Stepwise ===\n")
print(names(coef(model_step)))

cat("\n=== Variables Removed by Stepwise ===\n")
removed <- setdiff(names(coef(model_multi)), names(coef(model_step)))
print(removed)

cat("\nAIC after stepwise:", round(AIC(model_step), 2), "\n")
```



```{r significance-check}
coef_table    <- summary(model_multi)$coefficients
coef_df       <- as.data.frame(coef_table)
coef_df$Predictor <- rownames(coef_df)
colnames(coef_df) <- c("Estimate", "SE", "z_value", "p_value", "Predictor")
coef_df           <- coef_df[order(coef_df$p_value), ]
coef_df$Significant <- ifelse(coef_df$p_value < 0.05, "Yes", "No")

cat("=== Non-significant predictors (p > 0.05) ===\n")
print(data.frame(
  Predictor = coef_df$Predictor[coef_df$Significant == "No"],
  p_value   = round(coef_df$p_value[coef_df$Significant == "No"], 6)
))
```

```{r anova-deviance}
anova(model_multi, test = "Chisq")
```

---

## Final Model — Equation, Coefficients & Odds Ratios

```{r final-model}
model_final <- model_step

cat("=== Final Model Summary ===\n")
summary(model_final)

mcfadden_final <- 1 - (model_final$deviance / model_final$null.deviance)
cat("\nFinal Model AIC:", round(AIC(model_final), 2), "\n")
cat("McFadden R²:    ", round(mcfadden_final, 4), "\n")
```

```{r final-model-equation}
cat("=== FINAL MODEL EQUATION ===\n\n")

coef_final <- coef(model_final)

cat("log( p / (1-p) ) = ",
    round(coef_final[1], 4), "\n", sep = "")

for(i in 2:length(coef_final)) {

  coef_value <- round(coef_final[i], 6)
  coef_name  <- names(coef_final)[i]

  if(coef_value >= 0){
    cat(" + ", coef_value, " * ", coef_name, "\n", sep = "")
  } else {
    cat(" - ", abs(coef_value), " * ", coef_name, "\n", sep = "")
  }
}

cat("\nWhere p = probability of subscribing to a term deposit\n")
```

```{r final-model-or-table}
or_final <- exp(cbind(OR = coef(model_final), confint(model_final)))

or_table <- data.frame(
  Predictor = rownames(or_final),
  OR        = round(or_final[, "OR"],     3),
  CI_lower  = round(or_final[, "2.5 %"],  3),
  CI_upper  = round(or_final[, "97.5 %"], 3),
  p_value   = round(summary(model_final)$coefficients[, 4], 4),
  row.names = NULL
)

or_table$Sig <- ifelse(or_table$p_value < 0.001, "***",
                ifelse(or_table$p_value < 0.01,  "**",
                ifelse(or_table$p_value < 0.05,  "*",
                ifelse(or_table$p_value < 0.10,  ".", ""))))

or_table <- or_table[-1, ]
rownames(or_table) <- NULL
print(or_table)
```

```{r final-model-interpretations}
cat("=== KEY COEFFICIENT INTERPRETATIONS ===\n\n")

cat("NUMERIC PREDICTORS:\n")
cat("duration:  OR =", round(exp(coef(model_final)["duration"]), 4),
    "— each extra second increases odds by",
    round((exp(coef(model_final)["duration"]) - 1) * 100, 2), "%\n")
cat("           Per extra minute: OR =",
    round(exp(60 * coef(model_final)["duration"]), 3),
    "— odds increase by",
    round((exp(60 * coef(model_final)["duration"]) - 1) * 100, 1), "%\n\n")
cat("age:       OR =", round(exp(coef(model_final)["age"]), 4),
    "— each extra year increases odds by",
    round((exp(coef(model_final)["age"]) - 1) * 100, 2), "%\n\n")
cat("balance:   Per €1000 increase: OR =",
    round(exp(1000 * coef(model_final)["balance"]), 3),
    "— increases odds by 3.2%\n\n")
cat("campaign:  OR =", round(exp(coef(model_final)["campaign"]), 4),
    "— each extra contact during campaign DECREASES odds by",
    round((1 - exp(coef(model_final)["campaign"])) * 100, 1), "%\n\n")

cat("JOB (reference = admin):\n")
cat("jobstudent:     OR =", round(exp(coef(model_final)["jobstudent"]), 3),
    "— students have highest odds relative to admin\n")
cat("jobretired:     OR =", round(exp(coef(model_final)["jobretired"]), 3),
    "— retired clients have elevated odds relative to admin\n")
cat("jobblue-collar: OR =", round(exp(coef(model_final)["jobblue-collar"]), 3),
    "— blue-collar workers have lower odds than admin\n\n")

cat("CONTACT (reference = cellular):\n")
cat("contactunknown: OR =", round(exp(coef(model_final)["contactunknown"]), 3),
    "— unknown contact type has much lower odds than cellular\n\n")

cat("PREVIOUS OUTCOME (reference = failure):\n")
cat("poutcomesuccess: OR =", round(exp(coef(model_final)["poutcomesuccess"]), 3),
    "— previous success dramatically increases odds of subscription\n")
```

> **Key findings from the final model:**
>
> - **`poutcome = success`** is the single strongest predictor (OR ≈ 13.562). A client whose previous campaign outcome was a success is nearly 13× more likely to subscribe again.
> - **`job = student`** (OR ≈ 2.27) and **`job = retired`** (OR ≈ 1.72) are the job categories most positively associated with subscription relative to the admin reference.
> - **`contact = unknown`** (OR ≈ 0.25) dramatically reduces the probability of subscription compared with cellular contact.
> - **`duration`**: each additional minute of call duration increases odds by approximately 26.6% (after adjusting for all other predictors).
> - **`campaign`**: more contacts during the current campaign slightly reduces odds (OR ≈ 0.91 per additional contact).

---

# Class Imbalance — Weighting & Threshold Tuning

## Class Distribution

```{r class-distribution}
cat("=== Class Distribution in Training Data ===\n")
table(train_data$y)
round(prop.table(table(train_data$y)) * 100, 2)
```

## Inverse Frequency Weighting

```{r class-weights}
n_no  <- sum(train_data$y == "no")
n_yes <- sum(train_data$y == "yes")
total <- nrow(train_data)

cat("Weight for no: ",  round(total / (2 * n_no),  4), "\n")
cat("Weight for yes:", round(total / (2 * n_yes), 4), "\n")

train_weights <- ifelse(train_data$y == "yes",
                        total / (2 * n_yes),
                        total / (2 * n_no))

model_weighted <- glm(formula(model_final),
                      data    = train_data,
                      family  = binomial,
                      weights = train_weights)

summary(model_weighted)
```

## Threshold Tuning

```{r threshold-tuning}
pred_valid <- predict(
  model_final,
  newdata = valid_data,
  type = "response"
)

thresholds <- c(0.5, 0.4, 0.3, 0.2)

cat("=== Threshold Tuning — Validation Data ===\n")
cat(sprintf("%-12s %-12s %-12s %-12s\n",
            "Threshold", "Sensitivity", "Specificity", "Accuracy"))

for (t in thresholds) {

  pred_t <- factor(
    ifelse(pred_valid >= t, "yes", "no"),
    levels = c("no", "yes")
  )

  cm_t <- table(
    Predicted = pred_t,
    Actual    = valid_data$y
  )

  sensitivity <- cm_t["yes", "yes"] / sum(cm_t[, "yes"])
  specificity <- cm_t["no",  "no"]  / sum(cm_t[, "no"])
  accuracy    <- sum(diag(cm_t)) / sum(cm_t)

  cat(sprintf(
    "%-12.1f %-12.3f %-12.3f %-12.3f\n",
    t, sensitivity, specificity, accuracy
  ))
}
```
```{r pridictor-test-set}
pred_test <- predict(
  model_final,
  newdata = test_data,
  type = "response"
)

pred_weighted_test <- predict(
  model_weighted,
  newdata = test_data,
  type = "response"
)

pred_weighted_valid <- predict(
  model_weighted,
  newdata = valid_data,
  type = "response"
)

```

```{r threshold-distribution-plot}
pred_df <- data.frame(probability = pred_test, actual = test_data$y)

ggplot(pred_df, aes(x = probability, fill = actual)) +
  geom_histogram(bins = 50, alpha = 0.6, position = "identity") +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "black", linewidth = 0.8) +
  geom_vline(xintercept = 0.2, linetype = "dashed", color = "red",   linewidth = 0.8) +
  annotate("text", x = 0.52, y = 500, label = "0.5", size = 3.5) +
  annotate("text", x = 0.22, y = 500, label = "0.2", size = 3.5, color = "red") +
  labs(title = "Predicted Probabilities by Actual Outcome",
       subtitle = "Black dashed = 0.5 threshold | Red dashed = 0.2 threshold",
       x = "Predicted Probability of Subscribing", y = "Count", fill = "Actual") +
  theme_minimal()
```

> Lowering the threshold from 0.5 to 0.2 increases sensitivity (ability to detect true
> subscribers) at the cost of reduced specificity. A threshold of 0.2 was selected as
> the best balance for the unweighted model.

---

# Threshold Comparison — Detailed Analysis

## Why Threshold Choice Matters

In logistic regression, the model outputs a **probability** (0 to 1) rather than a
direct class label. A **threshold** (also called a cut-off) is needed to convert that
probability into a predicted class: values at or above the threshold are predicted as
"yes", values below as "no".

The default threshold of 0.5 assumes equal cost for false positives and false negatives.
In an imbalanced dataset like this one (only ~11.7% are "yes"), this default tends to
under-predict the minority class because the model is rarely confident enough to exceed
0.5 for "yes" cases.

Lowering the threshold increases **Sensitivity** (the model catches more real "yes"
cases) but reduces **Specificity** (more "no" clients are incorrectly predicted as "yes").
The optimal threshold depends on the business goal:

- If the **cost of missing a subscriber** is high → prioritise Sensitivity → lower threshold
- If the **cost of contacting a non-subscriber** is high → prioritise Specificity → raise threshold

## Full Threshold Grid

```{r threshold-grid}
thresholds_full <- seq(0.1, 0.9, by = 0.05)

threshold_results <- data.frame(
  Threshold   = numeric(),
  Sensitivity = numeric(),
  Specificity = numeric(),
  Accuracy    = numeric(),
  Kappa       = numeric(),
  F1          = numeric()
)

for (t in thresholds_full) {
  pred_t  <- factor(ifelse(pred_valid >= t, "yes", "no"), levels = c("no", "yes"))
  cm_t    <- confusionMatrix(pred_t, valid_data$y, positive = "yes")
  tp      <- cm_t$table["yes", "yes"]
  fp      <- cm_t$table["yes", "no"]
  fn      <- cm_t$table["no",  "yes"]
  prec    <- ifelse((tp + fp) == 0, 0, tp / (tp + fp))
  rec     <- cm_t$byClass["Sensitivity"]
  f1      <- ifelse((prec + rec) == 0, 0, 2 * prec * rec / (prec + rec))

  threshold_results <- rbind(threshold_results, data.frame(
    Threshold   = t,
    Sensitivity = round(cm_t$byClass["Sensitivity"], 3),
    Specificity = round(cm_t$byClass["Specificity"], 3),
    Accuracy    = round(cm_t$overall["Accuracy"],    3),
    Kappa       = round(cm_t$overall["Kappa"],       3),
    F1          = round(f1, 3)
  ))
}

rownames(threshold_results) <- NULL
print(threshold_results)
```

```{r threshold-tradeoff-plot}
thresh_long <- threshold_results %>%
  pivot_longer(cols = c(Sensitivity, Specificity, Accuracy, F1, Kappa),
               names_to = "Metric", values_to = "Value")

ggplot(thresh_long, aes(x = Threshold, y = Value, colour = Metric)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = 0.2, linetype = "dashed", colour = "black") +
  geom_vline(xintercept = 0.5, linetype = "dotted", colour = "grey40") +
  annotate("text", x = 0.21, y = 0.15, label = "Chosen\n(0.2)",
           size = 3, hjust = 0) +
  annotate("text", x = 0.51, y = 0.15, label = "Default\n(0.5)",
           size = 3, hjust = 0, colour = "grey40") +
  scale_colour_brewer(palette = "Set1") +
  labs(title    = "Sensitivity / Specificity / Accuracy / F1 / Kappa by Threshold",
subtitle = "Unweighted model on validation data",
x = "Classification Threshold", y = "Metric Value") +
  theme_minimal()
```

```{r threshold-sensitivity-specificity}
ggplot(threshold_results, aes(x = Threshold)) +
  geom_line(aes(y = Sensitivity, colour = "Sensitivity"), linewidth = 1.2) +
  geom_line(aes(y = Specificity, colour = "Specificity"), linewidth = 1.2) +
  geom_line(aes(y = F1,          colour = "F1 Score"),    linewidth = 1.2) +
  geom_vline(xintercept = 0.2, linetype = "dashed") +
  scale_colour_manual(values = c("Sensitivity" = "tomato",
                                 "Specificity" = "steelblue",
                                 "F1 Score"    = "darkgreen")) +
  labs(title  = "Sensitivity–Specificity Trade-off by Threshold",
       x = "Threshold", y = "Value", colour = "Metric") +
  theme_minimal()
```

## Why Threshold 0.2 Was Chosen

```{r threshold-summary-table}
key_thresholds <- threshold_results[threshold_results$Threshold %in% c(0.2, 0.3, 0.4, 0.5, 0.6), ]
rownames(key_thresholds) <- NULL
print(key_thresholds)
```

> | Threshold | Effect |
> |-----------|--------|
> | **0.5 (default)** | High accuracy and specificity, but very low sensitivity — misses most actual subscribers |
> | **0.2 (chosen)** | Best balance: highest Kappa (0.477) and F1 (0.545), catches 61.2% of subscribers, specificity remains acceptable at 91.4% |
> | **0.1** | Sensitivity rises further but specificity drops sharply — too many non-subscribers incorrectly flagged |
>
> A threshold of **0.2** was selected because it maximises both Kappa (0.477) and
> F1 score (0.545) — the two metrics most robust to class imbalance. Accuracy alone
> is misleading here: a model that predicts "no" for every client would still achieve
> 88.3% accuracy due to the class imbalance. Kappa corrects for this by measuring
> agreement beyond chance. In a marketing context, threshold 0.2 is the most
> commercially sensible choice: it identifies over 60% of genuine subscribers while
> keeping the false positive rate low enough to avoid wasting campaign resources.

---

## BSMOTE — Borderline Synthetic Minority Oversampling

Borderline-SMOTE (BSMOTE) is a third approach to class imbalance. Unlike threshold
tuning (which changes only the decision boundary) or class weighting (which adjusts
the loss function), BSMOTE **modifies the training data itself** by synthesising new
"yes" observations near the decision boundary — the area where the two classes are
hardest to distinguish.

The logic: only the "borderline" minority observations — those surrounded by majority
class neighbours — are used as seeds for synthesis. New synthetic points are generated
along the line between a borderline minority observation and one of its nearest
minority neighbours. This makes the model more sensitive in precisely the region where
it struggles most.

````{r bsmote-setup}
# install.packages("smotefamily")   # run once if not installed
library(smotefamily)
````

````{r bsmote-prep}
# BSMOTE requires a numeric matrix — convert factors to dummies first
train_numeric <- model.matrix(
  y ~ duration + age + balance + campaign +
      job + marital + education + contact + poutcome,
  data = train_data
)[, -1]   # remove intercept column

train_outcome <- as.numeric(train_data$y == "yes")   # 1 = yes, 0 = no

cat("=== Class Distribution BEFORE BSMOTE ===\n")
table(train_data$y)
round(prop.table(table(train_data$y)) * 100, 2)
````

````{r bsmote-run}
set.seed(123)

bsmote_result <- BLSMOTE(
  X       = as.data.frame(train_numeric),
  target  = train_outcome,
  K       = 5,    # nearest neighbours within minority class
  C       = 5,    # nearest neighbours used to identify borderline cases
  dupSize = 0     # auto-determine duplication size to balance classes
)

# Extract balanced dataset and reconstruct outcome factor
train_bsmote       <- bsmote_result$data
train_bsmote$y     <- as.factor(ifelse(train_bsmote$class == 1, "yes", "no"))
train_bsmote$y     <- relevel(train_bsmote$y, ref = "no")
train_bsmote$class <- NULL

cat("=== Class Distribution AFTER BSMOTE ===\n")
table(train_bsmote$y)
round(prop.table(table(train_bsmote$y)) * 100, 2)
````

> BSMOTE increases the number of "yes" observations in the training set by
> generating synthetic cases near the class boundary. The test set is **never
> resampled** — evaluation always uses the original imbalanced test data so
> that results reflect real-world conditions.

> BSMOTE requires conversion of categorical variables to dummy variables before oversampling. Therefore the logistic regression fitted on the BSMOTE dataset uses the dummy-variable representation of the predictors rather than the original factor variables.

````{r bsmote-model}
# Refit logistic regression on the BSMOTE-balanced training data
model_bsmote <- glm(
  y ~ .,
  data   = train_bsmote,
  family = binomial
)

cat("BSMOTE model AIC:", round(AIC(model_bsmote), 2), "\n")
summary(model_bsmote)
````

````{r bsmote-predict}
# Predict on the ORIGINAL (imbalanced) test set
# Must rebuild the same dummy matrix used during training
X_test_bsmote <- as.data.frame(
  model.matrix(
    y ~ duration + age + balance + campaign +
        job + marital + education + contact + poutcome,
    data = test_data
  )[, -1]
)

pred_bsmote_prob  <- predict(model_bsmote,
                              newdata = X_test_bsmote,
                              type    = "response")

pred_bsmote_class <- factor(
  ifelse(pred_bsmote_prob >= 0.5, "yes", "no"),
  levels = c("no", "yes")
)

cat("=== Confusion Matrix — BSMOTE Model (Threshold 0.5) ===\n")
cm_bsmote <- confusionMatrix(pred_bsmote_class, test_data$y, positive = "yes")
print(cm_bsmote)
````

````{r bsmote-roc}
roc_bsmote <- roc(test_data$y, pred_bsmote_prob,
                  levels = c("no", "yes"), quiet = TRUE)

cat("BSMOTE model AUC:", round(auc(roc_bsmote), 4), "\n")
````

> BSMOTE typically yields higher sensitivity than the default threshold of 0.5
> on the unweighted model, because the model was trained on a more balanced
> dataset and learned a less conservative decision boundary. However AUC tends
> to remain similar across all three methods — AUC measures ranking ability,
> which is less affected by class rebalancing than threshold-dependent metrics.




# Regularisation for Variable Selection

Regularisation methods shrink or eliminate coefficients to reduce overfitting and
perform automatic variable selection. All three standard approaches are applied here
— LASSO (L1), Ridge (L2), and Elastic Net — and compared against the stepwise
AIC final model.

## Prepare Data for glmnet


```{r reg-setup}
library(glmnet)

# glmnet requires a numeric model matrix (no intercept column — glmnet adds it)
X_train <- model.matrix(y ~ duration + age + balance + campaign +
                          job + marital + education + contact + poutcome,
                        data = train_data)[, -1]

X_test  <- model.matrix(y ~ duration + age + balance + campaign +
                          job + marital + education + contact + poutcome,
                        data = test_data)[, -1]

# Binary outcome must be numeric 0/1 for glmnet
y_train <- as.numeric(train_data$y == "yes")
y_test  <- as.numeric(test_data$y  == "yes")
```

> `model.matrix()` converts all factor variables into dummy variables automatically,
> which is required by `glmnet`. The intercept column is removed because `glmnet`
> adds its own.

---

## LASSO (L1 Regularisation)

LASSO (Least Absolute Shrinkage and Selection Operator) applies an L1 penalty that
can shrink some coefficients exactly to zero, effectively removing variables from
the model. This makes it a direct variable selection tool.

```{r lasso-cv}
set.seed(123)
cv_lasso <- cv.glmnet(X_train, y_train,
                      family  = "binomial",
                      alpha   = 1,          # alpha=1 => LASSO
                      nfolds  = 10)

plot(cv_lasso)
title("LASSO: Cross-Validated Deviance vs log(lambda)", line = 2.5)
```

```{r lasso-lambda}
cat("=== LASSO Optimal Lambda ===\n")
cat("lambda.min (lowest CV error):    ", round(cv_lasso$lambda.min,    6), "\n")
cat("lambda.1se (most regularised within 1 SE):", round(cv_lasso$lambda.1se, 6), "\n")
```

```{r lasso-coefficients}
cat("=== LASSO Coefficients at lambda.min ===\n")
lasso_coef_min <- coef(cv_lasso, s = "lambda.min")
print(lasso_coef_min)

cat("\n=== LASSO Coefficients at lambda.1se (more regularised) ===\n")
lasso_coef_1se <- coef(cv_lasso, s = "lambda.1se")
print(lasso_coef_1se)

# Variables shrunk to zero at lambda.1se
zero_vars <- rownames(lasso_coef_1se)[which(lasso_coef_1se[, 1] == 0)]
cat("\nVariables eliminated by LASSO (lambda.1se):\n")
print(zero_vars)
```

```{r lasso-predictions}
pred_lasso <- predict(cv_lasso, newx = X_test,
                      s    = "lambda.min",
                      type = "response")[, 1]

roc_lasso  <- roc(test_data$y, pred_lasso, levels = c("no", "yes"), quiet = TRUE)
cat("LASSO AUC:", round(auc(roc_lasso), 4), "\n")
```

> LASSO performs both regularisation and variable selection simultaneously. Coefficients
> driven to zero at `lambda.1se` represent variables the model can drop without
> meaningfully increasing prediction error. Variables retained by LASSO are the most
> robustly associated with the outcome.

---

## Ridge (L2 Regularisation)

Ridge regression applies an L2 penalty that shrinks all coefficients towards zero
but does **not** eliminate any variable entirely. It is best used when all predictors
are expected to contribute at least a little, and is particularly useful when
multicollinearity is present.

```{r ridge-cv}
set.seed(123)
cv_ridge <- cv.glmnet(X_train, y_train,
                      family = "binomial",
                      alpha  = 0,           # alpha=0 => Ridge
                      nfolds = 10)

plot(cv_ridge)
title("Ridge: Cross-Validated Deviance vs log(lambda)", line = 2.5)
```

```{r ridge-coefficients}
cat("=== Ridge Coefficients at lambda.min ===\n")
ridge_coef <- coef(cv_ridge, s = "lambda.min")
print(ridge_coef)
```

```{r ridge-predictions}
pred_ridge <- predict(cv_ridge, newx = X_test,
                      s    = "lambda.min",
                      type = "response")[, 1]

roc_ridge  <- roc(test_data$y, pred_ridge, levels = c("no", "yes"), quiet = TRUE)
cat("Ridge AUC:", round(auc(roc_ridge), 4), "\n")
```

> Unlike LASSO, Ridge retains all variables but applies proportional shrinkage. It
> tends to perform well when many small effects are present. Because it cannot zero
> out coefficients, it is less interpretable as a selection tool but can yield better
> predictive accuracy.

---

## Elastic Net (L1 + L2 Combined)

Elastic Net combines both penalties, using a mixing parameter `alpha` between 0 and 1.
Here `alpha = 0.5` is used, giving equal weight to LASSO and Ridge penalties.

```{r enet-cv}
set.seed(123)
cv_enet <- cv.glmnet(X_train, y_train,
                     family  = "binomial",
                     alpha   = 0.5,         # 50% LASSO, 50% Ridge
                     nfolds  = 10)

plot(cv_enet)
title("Elastic Net: Cross-Validated Deviance vs log(lambda)", line = 2.5)
```

```{r enet-coefficients}
cat("=== Elastic Net Coefficients at lambda.min ===\n")
enet_coef <- coef(cv_enet, s = "lambda.min")
print(enet_coef)

zero_enet <- rownames(enet_coef)[which(enet_coef[, 1] == 0)]
cat("\nVariables eliminated by Elastic Net (lambda.min):\n")
print(zero_enet)
```

```{r enet-predictions}
pred_enet <- predict(cv_enet, newx = X_test,
                     s    = "lambda.min",
                     type = "response")[, 1]

roc_enet  <- roc(test_data$y, pred_enet, levels = c("no", "yes"), quiet = TRUE)
cat("Elastic Net AUC:", round(auc(roc_enet), 4), "\n")
```

> Elastic Net is a useful compromise: it can zero out irrelevant variables like LASSO
> while handling correlated predictors more stably like Ridge. It is generally
> preferred over pure LASSO when groups of correlated variables are present.

---

## Regularisation Comparison — All Methods

```{r reg-comparison}
# AUC for the stepwise final model (already computed earlier)
roc_final_reg <- roc(test_data$y, pred_test,
                     levels = c("no", "yes"), quiet = TRUE)

reg_comparison <- data.frame(
  Method        = c("Stepwise AIC (Final Model)",
                    "LASSO (lambda.min)",
                    "Ridge (lambda.min)",
                    "Elastic Net (alpha=0.5)"),
  AUC           = round(c(auc(roc_final_reg),
                           auc(roc_lasso),
                           auc(roc_ridge),
                           auc(roc_enet)), 4),
  Variables_Kept = c("9 predictors",
                      paste(sum(lasso_coef_min[-1] != 0), "dummy columns"),
                      "All (shrunk)",
                      paste(sum(enet_coef[-1] != 0),  "dummy columns"))
)

print(reg_comparison, row.names = FALSE)
```

```{r reg-roc-comparison}
# Plot all ROC curves together
plot(roc_final_reg, col = "steelblue",  lwd = 2,
     main = "ROC Comparison — Stepwise vs Regularisation Methods")
lines(roc_lasso,   col = "tomato",      lwd = 2, lty = 2)
lines(roc_ridge,   col = "darkgreen",   lwd = 2, lty = 3)
lines(roc_enet,    col = "purple",      lwd = 2, lty = 4)
abline(a = 0, b = 1, lty = 2, col = "grey60")

legend("bottomright",
       legend = c(paste("Stepwise AIC   AUC =", round(auc(roc_final_reg), 3)),
                  paste("LASSO          AUC =", round(auc(roc_lasso),     3)),
                  paste("Ridge          AUC =", round(auc(roc_ridge),     3)),
                  paste("Elastic Net    AUC =", round(auc(roc_enet),      3))),
       col = c("steelblue", "tomato", "darkgreen", "purple"),
       lwd = 2, lty = 1:4)
```

> All four methods produce very similar AUC values, which confirms that the stepwise
> model's variable selection was robust. If LASSO reaches a comparable AUC with fewer
> non-zero dummy columns, it provides a simpler and equally valid model. The small
> differences in AUC across methods suggest the predictive signal in this dataset is
> strong and consistent regardless of the selection approach used.

---

# Model Evaluation

## ROC Curves & AUC

```{r roc-curves}
pred_simple <- predict(model_duration, newdata = test_data, type = "response")

roc_simple   <- roc(test_data$y, pred_simple,      levels = c("no", "yes"), quiet = TRUE)
roc_final    <- roc(test_data$y, pred_test,  levels = c("no", "yes"), quiet = TRUE)
roc_weighted <- roc(test_data$y, pred_weighted_test,    levels = c("no", "yes"), quiet = TRUE)

cat("=== AUC Comparison ===\n")
cat("Simple model   AUC:", round(auc(roc_simple),   4), "\n")
cat("Final model    AUC:", round(auc(roc_final),    4), "\n")
cat("Weighted model AUC:", round(auc(roc_weighted), 4), "\n")

plot(roc_final,   col = "steelblue",  lwd = 2, main = "ROC Curve Comparison")
lines(roc_simple, col = "tomato",     lwd = 2, lty = 2)
lines(roc_weighted, col = "darkgreen", lwd = 2, lty = 3)
abline(a = 0, b = 1, lty = 2, col = "grey60")
legend("bottomright",
       legend = c(paste("Final model AUC    =", round(auc(roc_final),    3)),
                  paste("Simple model AUC   =", round(auc(roc_simple),   3)),
                  paste("Weighted model AUC =", round(auc(roc_weighted), 3)),
                  "Random classifier"),
       col = c("steelblue", "tomato", "darkgreen", "grey60"),
       lwd = 2, lty = c(1, 2, 3, 2))
```

> The final unweighted model (AUC ≈ 0.885) substantially outperforms the simple
> duration-only model (AUC ≈ 0.807), confirming that the additional predictors
> contribute meaningfully to predictive power.

## Confusion Matrices — Three Scenarios

```{r confusion-matrices}
pred_class_05 <- factor(ifelse(pred_test >= 0.5, "yes", "no"), levels = c("no", "yes"))
pred_class_02 <- factor(ifelse(pred_test >= 0.2, "yes", "no"), levels = c("no", "yes"))
pred_class_wt <- factor(ifelse(pred_weighted_test   >= 0.5, "yes", "no"), levels = c("no", "yes"))

cat("=== Confusion Matrix: Unweighted, Threshold 0.5 ===\n")
cm1 <- confusionMatrix(pred_class_05, test_data$y, positive = "yes")
print(cm1)

cat("\n=== Confusion Matrix: Unweighted, Threshold 0.2 ===\n")
cm2 <- confusionMatrix(pred_class_02, test_data$y, positive = "yes")
print(cm2)

cat("\n=== Confusion Matrix: Weighted, Threshold 0.5 ===\n")
cm3 <- confusionMatrix(pred_class_wt, test_data$y, positive = "yes")
print(cm3)
```

```{r confusion-matrix-plots}
plot_cm <- function(cm, title) {
  cm_df <- as.data.frame(cm$table)
  ggplot(cm_df, aes(x = Reference, y = Prediction, fill = Freq)) +
    geom_tile(color = "white") +
    geom_text(aes(label = Freq), size = 8, fontface = "bold") +
    scale_fill_gradient(low = "white", high = "steelblue") +
    labs(title = title, x = "Actual", y = "Predicted") +
    theme_minimal() +
    theme(legend.position = "none",
          axis.text  = element_text(size = 12),
          axis.title = element_text(size = 13),
          plot.title = element_text(size = 14, face = "bold"))
}

plot_cm(cm1, "Confusion Matrix — Unweighted, Threshold 0.5")
plot_cm(cm2, "Confusion Matrix — Unweighted, Threshold 0.2")
plot_cm(cm3, "Confusion Matrix — Weighted,   Threshold 0.5")
```

## Performance Metric Comparison

```{r metrics-comparison}
metrics_df <- data.frame(
  Scenario    = c("Unweighted t=0.5", "Unweighted t=0.2", "Weighted   t=0.5"),
  Accuracy    = c(round(cm1$overall["Accuracy"],    3),
                  round(cm2$overall["Accuracy"],    3),
                  round(cm3$overall["Accuracy"],    3)),
  Sensitivity = c(round(cm1$byClass["Sensitivity"], 3),
                  round(cm2$byClass["Sensitivity"], 3),
                  round(cm3$byClass["Sensitivity"], 3)),
  Specificity = c(round(cm1$byClass["Specificity"], 3),
                  round(cm2$byClass["Specificity"], 3),
                  round(cm3$byClass["Specificity"], 3)),
  Kappa       = c(round(cm1$overall["Kappa"],       3),
                  round(cm2$overall["Kappa"],       3),
                  round(cm3$overall["Kappa"],       3))
)

print(metrics_df, row.names = FALSE)
```

> The **unweighted model with threshold = 0.2** achieves the best overall balance:
> Sensitivity = 60.9%, Specificity = 91.6%, Accuracy = 88.1%, Kappa = 0.471. This
> scenario best handles the class imbalance while maintaining acceptable accuracy.

## Predicted Probability Distribution

```{r prob-distribution}
pred_df <- data.frame(probability = pred_test, actual = test_data$y)

ggplot(pred_df, aes(x = probability, fill = actual)) +
  geom_histogram(bins = 50, alpha = 0.6, position = "identity") +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  geom_vline(xintercept = 0.2, linetype = "dashed",
             color = "black", linewidth = 0.8) +
  annotate("text", x = 0.22, y = 400,
           label = "Chosen threshold = 0.2", size = 3.5, hjust = 0) +
  labs(title = "Predicted Probabilities by Actual Outcome",
       x = "Predicted Probability of Subscribing",
       y = "Count", fill = "Actual") +
  theme_minimal()
```



---

## Predictions on Training vs Test Data

Comparing model performance on training and test data is essential for detecting
**overfitting** — a situation where the model learns the training data too well but
fails to generalise.

### Predicted Probabilities: Training vs Test

```{r train-test-predictions}
# Predict on training data
pred_train <- predict(model_final, newdata = train_data, type = "response")

# Apply chosen threshold of 0.2
class_train_02 <- factor(ifelse(pred_train      >= 0.2, "yes", "no"),
                          levels = c("no", "yes"))
class_test_02  <- factor(ifelse(pred_test >= 0.2, "yes", "no"),
                          levels = c("no", "yes"))

cat("=== Training Data Predictions (Threshold = 0.2) ===\n")
cm_train <- confusionMatrix(class_train_02,
                             factor(train_data$y, levels = c("no", "yes")),
                             positive = "yes")
print(cm_train)

cat("\n=== Test Data Predictions (Threshold = 0.2) ===\n")
cm_test <- confusionMatrix(class_test_02, test_data$y, positive = "yes")
print(cm_test)
```


### Side-by-Side Metric Comparison

```{r train-test-metric-comparison}
train_test_metrics <- data.frame(
  Metric      = c("Accuracy", "Sensitivity", "Specificity", "Kappa"),
  Training    = c(round(cm_train$overall["Accuracy"],    3),
                  round(cm_train$byClass["Sensitivity"], 3),
                  round(cm_train$byClass["Specificity"], 3),
                  round(cm_train$overall["Kappa"],       3)),
  Test        = c(round(cm_test$overall["Accuracy"],    3),
                  round(cm_test$byClass["Sensitivity"], 3),
                  round(cm_test$byClass["Specificity"], 3),
                  round(cm_test$overall["Kappa"],       3))
)

print(train_test_metrics, row.names = FALSE)
```

```{r train-test-bar-plot}
train_test_long <- train_test_metrics %>%
  pivot_longer(cols = c(Training, Test),
               names_to = "Split", values_to = "Value")

ggplot(train_test_long, aes(x = Metric, y = Value, fill = Split)) +
  geom_col(position = "dodge") +
  geom_text(aes(label = Value),
            position = position_dodge(width = 0.9),
            vjust = -0.4, size = 3.5) +
  scale_fill_manual(values = c("Training" = "steelblue", "Test" = "tomato")) +
  labs(title = "Training vs Test Performance (Threshold = 0.2)",
       x = NULL, y = "Metric Value") +
  theme_minimal() +
  ylim(0, 1.05)
```

### Predicted Probability Distributions: Train vs Test

```{r prob-dist-train-test}
prob_compare <- data.frame(
  probability = c(pred_train, pred_test),
  actual      = c(as.character(train_data$y),
                  as.character(test_data$y)),
  split       = c(rep("Training", length(pred_train)),
                  rep("Test", length(pred_test)))
)

ggplot(prob_compare, aes(x = probability, fill = actual)) +
  geom_histogram(bins = 50, alpha = 0.6, position = "identity") +
  facet_wrap(~ split) +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  geom_vline(xintercept = 0.2, linetype = "dashed", colour = "black") +
  labs(title = "Predicted Probability Distributions — Training vs Test",
       x = "Predicted Probability of Subscribing",
       y = "Count", fill = "Actual") +
  theme_minimal()
```

### ROC: Training vs Test

```{r roc-train-vs-test}
roc_train <- roc(train_data$y, pred_train,
                 levels = c("no", "yes"), quiet = TRUE)
roc_test  <- roc(test_data$y,  pred_test,
                 levels = c("no", "yes"), quiet = TRUE)

plot(roc_train, col = "steelblue", lwd = 2,
     main = "ROC Curve — Training vs Test Data")
lines(roc_test, col = "tomato", lwd = 2, lty = 2)
abline(a = 0, b = 1, lty = 2, col = "grey60")

legend("bottomright",
       legend = c(paste("Training AUC =", round(auc(roc_train), 4)),
                  paste("Test AUC     =", round(auc(roc_test),  4))),
       col = c("steelblue", "tomato"), lwd = 2, lty = c(1, 2))

cat("Training AUC:", round(auc(roc_train), 4), "\n")
cat("Test AUC:    ", round(auc(roc_test),  4), "\n")
cat("Difference:  ", round(auc(roc_train) - auc(roc_test), 4), "\n")
```

> If training and test metrics are very close, the model generalises well and is not
> overfit. A large gap — particularly a much higher training AUC — would signal that
> the model has memorised the training data rather than learned the underlying pattern.
> Small differences (< 0.01–0.02 in AUC) are entirely normal and expected.

---

>>>>>



## Full Imbalance Handling Comparison

```{r imbalance-full-comparison}
comparison_imbalance <- data.frame(
  Method = c(
    "Unweighted  t=0.5  (baseline)",
    "Unweighted  t=0.2  (threshold tuned)",
    "Class weighted  t=0.5",
    "BSMOTE  t=0.5"
  ),
  AUC = c(
    round(auc(roc_final),    4),
    round(auc(roc_final),    4),
    round(auc(roc_weighted), 4),
    round(auc(roc_bsmote),   4)
  ),
  Sensitivity = c(
    round(cm1$byClass["Sensitivity"],       3),
    round(cm_test$byClass["Sensitivity"],   3),
    round(cm3$byClass["Sensitivity"],       3),
    round(cm_bsmote$byClass["Sensitivity"], 3)
  ),
  Specificity = c(
    round(cm1$byClass["Specificity"],       3),
    round(cm_test$byClass["Specificity"],   3),
    round(cm3$byClass["Specificity"],       3),
    round(cm_bsmote$byClass["Specificity"], 3)
  ),
  Accuracy = c(
    round(cm1$overall["Accuracy"],       3),
    round(cm_test$overall["Accuracy"],   3),
    round(cm3$overall["Accuracy"],       3),
    round(cm_bsmote$overall["Accuracy"], 3)
  ),
  Kappa = c(
    round(cm1$overall["Kappa"],       3),
    round(cm_test$overall["Kappa"],   3),
    round(cm3$overall["Kappa"],       3),
    round(cm_bsmote$overall["Kappa"], 3)
  )
)

print(comparison_imbalance, row.names = FALSE)
```

```{r imbalance-roc-comparison}
plot(roc_final,    col = "steelblue", lwd = 2,
     main = "ROC Comparison — Imbalance Handling Methods")
lines(roc_weighted, col = "purple",   lwd = 2, lty = 2)
lines(roc_bsmote,   col = "tomato",   lwd = 2, lty = 3)
abline(a = 0, b = 1, lty = 2, col = "grey60")

legend("bottomright",
       legend = c(
         paste("Unweighted t=0.2      AUC =", round(auc(roc_final),    3)),
         paste("Class weighted t=0.5  AUC =", round(auc(roc_weighted), 3)),
         paste("BSMOTE t=0.5          AUC =", round(auc(roc_bsmote),   3))
       ),
       col = c("steelblue", "purple", "tomato"),
       lwd = 2, lty = 1:3)
```


```{r imbalance-conclusion}
cat("=== IMBALANCE HANDLING — CONCLUSION ===\n\n")

best_kappa  <- comparison_imbalance$Method[which.max(comparison_imbalance$Kappa)]
best_sens   <- comparison_imbalance$Method[which.max(comparison_imbalance$Sensitivity)]

cat("Highest Kappa:      ", best_kappa, "\n")
cat("Highest Sensitivity:", best_sens,  "\n\n")

cat("Summary of approaches:\n")
cat("- Threshold t=0.2   : changes decision boundary only; training data unchanged\n")
cat("- Class weighting   : adjusts training loss; penalises minority misclassification\n")
cat("- BSMOTE            : synthesises new minority observations near the boundary\n\n")
cat("All three methods improve sensitivity over the default t=0.5 baseline.\n")
cat("Class weighting achieves the highest sensitivity at t=0.5 (0.773).\n")
cat("BSMOTE achieves comparable sensitivity (0.748) with slightly better specificity.\n")
cat("Both approaches substantially outperform the default unweighted threshold.\n")
```

> The table and ROC plot above give the definitive comparison across all four
> scenarios. The **unweighted model with threshold = 0.2** remains the recommended
> approach for this dataset: it achieves the best Kappa without modifying the
> training data, keeping the logistic regression coefficients and odds ratios
> directly interpretable.



>>>>>

---

## Predicted Subscription Probability: Actual vs Predicted

This section compares the **real observed subscription rate** against the **model's
predicted probabilities**, broken down by key variables. This helps verify that the
model is correctly calibrated and not systematically over- or under-predicting for
specific groups.

### Observed vs Predicted — By Job Type

```{r obs-pred-job}
compare_job <- test_data %>%
  mutate(pred_prob = pred_test) %>%
  group_by(job) %>%
  summarise(
    Observed_Rate  = round(mean(y == "yes") * 100, 2),
    Predicted_Rate = round(mean(pred_prob)  * 100, 2),
    n              = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(Observed_Rate))

print(compare_job)
```

```{r obs-pred-job-plot}
compare_job_long <- compare_job %>%
  pivot_longer(cols = c(Observed_Rate, Predicted_Rate),
               names_to = "Type", values_to = "Rate")

ggplot(compare_job_long, aes(x = reorder(job, Rate), y = Rate, fill = Type)) +
  geom_col(position = "dodge") +
  coord_flip() +
  scale_fill_manual(values = c("Observed_Rate"  = "steelblue",
                               "Predicted_Rate" = "tomato")) +
  labs(title = "Observed vs Predicted Subscription Rate by Job Type",
       x = "Job", y = "Subscription Rate (%)", fill = NULL) +
  theme_minimal()
```

### Observed vs Predicted — By Previous Campaign Outcome

```{r obs-pred-poutcome}
compare_pout <- test_data %>%
  mutate(pred_prob = pred_test) %>%
  group_by(poutcome) %>%
  summarise(
    Observed_Rate  = round(mean(y == "yes") * 100, 2),
    Predicted_Rate = round(mean(pred_prob)  * 100, 2),
    n              = n(),
    .groups = "drop"
  )

print(compare_pout)

compare_pout %>%
  pivot_longer(cols = c(Observed_Rate, Predicted_Rate),
               names_to = "Type", values_to = "Rate") %>%
  ggplot(aes(x = poutcome, y = Rate, fill = Type)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Observed_Rate"  = "steelblue",
                               "Predicted_Rate" = "tomato")) +
  labs(title = "Observed vs Predicted Subscription Rate by Previous Outcome",
       x = "Previous Outcome", y = "Subscription Rate (%)", fill = NULL) +
  theme_minimal()
```

### Observed vs Predicted — By Education Level

```{r obs-pred-education}
compare_edu <- test_data %>%
  mutate(pred_prob = pred_test) %>%
  group_by(education) %>%
  summarise(
    Observed_Rate  = round(mean(y == "yes") * 100, 2),
    Predicted_Rate = round(mean(pred_prob)  * 100, 2),
    n              = n(),
    .groups = "drop"
  )

print(compare_edu)

compare_edu %>%
  pivot_longer(cols = c(Observed_Rate, Predicted_Rate),
               names_to = "Type", values_to = "Rate") %>%
  ggplot(aes(x = education, y = Rate, fill = Type)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Observed_Rate"  = "steelblue",
                               "Predicted_Rate" = "tomato")) +
  labs(title = "Observed vs Predicted Subscription Rate by Education",
       x = "Education", y = "Subscription Rate (%)", fill = NULL) +
  theme_minimal()
```

> Across all three grouping variables, the predicted rates closely track the observed
> rates. This confirms that the model is well-calibrated — it does not systematically
> overestimate or underestimate the probability of subscription for any particular
> subgroup. Small discrepancies are expected and represent normal model error rather
> than systematic bias.

---

# Model Diagnostics

## Variance Inflation Factors (VIF)

```{r vif}
cat("=== Variance Inflation Factors (VIF) ===\n")
vif_vals <- vif(model_final)
print(round(vif_vals, 3))

vif_df <- as.data.frame(vif_vals)
cat("\nAny GVIF > 2.24 (equivalent to VIF > 5)?\n")
print(vif_df[vif_df[, 3] > 2.24, , drop = FALSE])
```

> All GVIF values are below 2.24, indicating **no multicollinearity** concerns in the final model.

## Cook's Distance — Influential Observations

```{r cooks-distance}
plot(model_final, which = 4,
     main = "Cook's Distance — Influential Observations")

cooksd <- cooks.distance(model_final)

cat("=== Top 10 Most Influential Observations ===\n")
print(round(sort(cooksd, decreasing = TRUE)[1:10], 6))

cat("\nObservations with Cook's D > 0.01:", sum(cooksd > 0.01), "\n")
cat("Observations with Cook's D > 0.1: ", sum(cooksd > 0.1),  "\n")
```

> The maximum Cook's Distance is approximately 0.007, well below common thresholds
> (0.01, 0.1, or 4/n). There are **no influential observations** that would substantially
> distort the model coefficients.

## Binned Residual Plot

```{r binned-residuals}
binnedplot(fitted(model_final),
           resid(model_final, type = "response"),
           nclass  = 40,
           main    = "Binned Residual Plot — Final Model",
           xlab    = "Fitted Probabilities",
           ylab    = "Average Residuals",
           col.int = "grey60")

cat("Points within grey bands = good model fit\n")
cat("Points outside bands = possible misfit or non-linearity\n")
```

> Most points fall within the grey 95% bands, indicating an overall good fit. A small
> number of points fall outside the bands at higher predicted probabilities, suggesting
> mild misfit in that region — acceptable for a model of this scale.

## Hosmer–Lemeshow Goodness-of-Fit Test

```{r hosmer-lemeshow}
hl_test <- hoslem.test(
  as.numeric(test_data$y == "yes"),
  pred_test,
  g = 10
)

print(hl_test)

cat("\n--- Interpretation ---\n")
cat("H0: model fits the data adequately\n")
cat("p > 0.05 = no evidence of poor fit\n")
cat("p < 0.05 = evidence of poor fit\n\n")
cat("Important: The Hosmer-Lemeshow test can be hypersensitive in large samples.\n")
cat("Even small deviations from perfect calibration may produce a significant p-value.\n")
cat("Therefore model fit should also be assessed using AUC, calibration plots and overall predictive performance.\n")
```

> The Hosmer–Lemeshow test returns p < 0.05; however, this is expected with a large
> sample (n > 45,000), where even trivial deviations are detected. The AUC of 0.882
> and McFadden R² of 0.295 confirm that the model fits well and is practically useful.

## Overall Model Fit Summary

```{r fit-summary}
cat("=== Overall Model Fit Summary ===\n\n")
cat("Null Deviance:    ", round(model_final$null.deviance, 2), "\n")
cat("Residual Deviance:", round(model_final$deviance, 2), "\n")
cat("Deviance Reduction:",
    round(model_final$null.deviance - model_final$deviance, 2), "\n\n")

mcfadden <- 1 - (model_final$deviance / model_final$null.deviance)
cat("McFadden R²:", round(mcfadden, 4), "\n")
cat("  0.10–0.20 = acceptable; 0.20–0.40 = good; >0.40 = excellent\n")
cat("  Our model:", round(mcfadden, 4), "= GOOD fit\n\n")

cat("AIC:", round(AIC(model_final), 2), "\n")
cat("BIC:", round(BIC(model_final), 2), "\n")
```

---

# K-Means Clustering & Customer Segmentation

## Why Clustering?

All analysis so far has been **supervised** — we used the subscription label `y` to
train the model. Clustering is **unsupervised** — it finds natural groupings in the data
without using the outcome label.

The goal here is **customer and campaign behaviour segmentation**: divide clients
into meaningful groups based on both their characteristics and their interaction
with the marketing campaign. The resulting segments are profiled to understand
different engagement patterns and subscription behaviour.

This is useful because:
- Different segments may need different marketing messages
- High-value segments can be prioritised in future campaigns
- It reveals structure in the data that logistic regression cannot show

## Prepare Data for Clustering
```{r}

```

```{r cluster-setup}
library(dplyr)
library(cluster)
library(factoextra)
```

```{r cluster-prep}
# Select numeric predictors relevant for segmentation
cluster_vars <- bank[, c("age", "balance", "duration", "campaign", "previous")]
cluster_vars <- na.omit(cluster_vars)

# Scale to mean=0, sd=1 so no single variable dominates due to magnitude
cluster_scaled <- scale(cluster_vars)

cat("Variables used for clustering:\n")
cat("age, balance, duration, campaign, previous\n\n")
cat("Rows:", nrow(cluster_scaled), "\n")
cat("All variables scaled to mean=0, sd=1\n")
```

> Scaling is essential for K-Means. Without it, `balance` (ranging from -8,000 to
> 100,000) would completely dominate `age` (18 to 95), making the cluster assignments
> driven almost entirely by balance alone.

## Finding the Optimal Number of Clusters

### Elbow Method

```{r cluster-elbow}
set.seed(123)
cluster_sample <- cluster_scaled[sample(nrow(cluster_scaled), 5000), ]

fviz_nbclust(
  cluster_sample,
  kmeans,
  method    = "wss",
  k.max     = 10,
  linecolor = "steelblue"
) +
  labs(title    = "Elbow Method — Within-Cluster Sum of Squares",
       subtitle = "Look for the 'elbow' where the curve bends sharply") +
  theme_minimal()
```

> The elbow method plots total within-cluster sum of squares (WSS) against the number
> of clusters. WSS decreases as k increases — but at some point the reduction slows
> dramatically. The optimal k is at the "elbow" — the point of diminishing returns
> where adding more clusters gives little improvement.

### Silhouette Method

```{r cluster-silhouette}
set.seed(123)
fviz_nbclust(
  cluster_sample,
  kmeans,
  method    = "silhouette",
  k.max     = 10,
  linecolor = "steelblue"
) +
  labs(title    = "Silhouette Method — Average Silhouette Width",
       subtitle = "Higher = better defined clusters. Peak = optimal k") +
  theme_minimal()
```

> The silhouette width measures how similar each observation is to its own cluster
> compared to the nearest neighbouring cluster. Values range from -1 to +1. A higher
> average silhouette width means the clusters are well-separated and internally compact.
> The k with the highest average silhouette width is optimal.

## Why k = 5? Justification from the Silhouette Plot

The silhouette method clearly showed k = 5 as the statistically optimal number of
clusters, with the highest average silhouette width (~0.33). While k = 3 is sometimes
chosen for simplicity, k = 5 is used here because:

- The silhouette plot peaks at k = 5 — this is the data-driven answer
- The improvement from k = 2 to k = 5 is substantial and consistent
- After k = 5 the silhouette drops sharply, confirming 5 is the true optimum
- For a final project, the statistically justified answer is more defensible

## Run K-Means with k = 5

```{r kmeans-fit}
set.seed(123)

# k = 5 selected based on silhouette method peak
km <- kmeans(
  cluster_scaled,
  centers  = 5,
  nstart   = 25,    # 25 random starts — keeps best solution
  iter.max = 100
)

cat("=== K-Means Results (k = 5) ===\n")
cat("Number of clusters:  ", nrow(km$centers), "\n")
cat("Cluster sizes:       ", km$size, "\n")
cat("Total WSS:           ", round(km$tot.withinss, 2), "\n")
cat("Between-cluster SS:  ", round(km$betweenss,    2), "\n")
cat("Variance explained:  ",
    round(km$betweenss / km$totss * 100, 1), "%\n")

# Add cluster labels back to the full dataset
bank$cluster <- as.factor(km$cluster)
```

> `nstart = 25` runs the algorithm 25 times with different random starting points
> and keeps the solution with the lowest total within-cluster sum of squares.
> This avoids getting stuck in a poor local optimum, which is especially important
> with k = 5.

## Sample for Visualisation

```{r cluster-sample}
set.seed(123)
sample_idx     <- sample(1:nrow(cluster_scaled), size = 5000)
cluster_sample <- cluster_scaled[sample_idx, ]

cat("Sample created:", nrow(cluster_sample), "rows\n")
```

## Visualise the Clusters

```{r cluster-viz}
# 5-colour palette — one per cluster
cluster_palette <- c("#2777AA", "#E03030", "#16A34A", "#D97706",
                     "#6D28D9")

# Use the sample for visualisation (full data too large for distance matrix)
fviz_cluster(
  list(data = cluster_sample, cluster = km$cluster[sample_idx]),
  geom         = "point",
  ellipse.type = "convex",
  palette      = cluster_palette,
  alpha        = 0.4,
  pointsize    = 0.8,
  main         = "K-Means Clusters (k=5) — Bank Clients (PCA projection)"
) +
  theme_minimal() +
  labs(subtitle = "Sample of 5,000 clients projected onto first two principal components")
```

> Because the data has 5 dimensions, `fviz_cluster` projects it onto the first two
> principal components for visualisation. Clusters that appear to overlap in this
> 2D view may be well-separated in the full 5-dimensional space.

## Cluster Profiles — Who is in Each Group?

```{r cluster-profiles}
# Summary statistics by cluster — using base R aggregate to avoid dplyr conflicts
cluster_profile <- aggregate(
  cbind(age, balance, duration, campaign) ~ cluster,
  data = bank,
  FUN  = mean
)

cluster_profile$n            <- as.vector(table(bank$cluster))
cluster_profile$pct_of_total <- round(cluster_profile$n / nrow(bank) * 100, 1)
cluster_profile$sub_rate_pct <- round(
  tapply(bank$y == "yes", bank$cluster, mean) * 100, 2)
cluster_profile$avg_age      <- round(cluster_profile$age,      1)
cluster_profile$avg_balance  <- round(cluster_profile$balance,  0)
cluster_profile$avg_duration <- round(cluster_profile$duration, 0)
cluster_profile$avg_campaign <- round(cluster_profile$campaign, 2)
cluster_profile$top_job <- sapply(
  levels(bank$cluster),
  function(cl) names(sort(table(bank$job[bank$cluster == cl]),
                           decreasing = TRUE))[1]
)

cluster_profile <- cluster_profile[, c("cluster", "n", "pct_of_total",
  "sub_rate_pct", "avg_age", "avg_balance", "avg_duration",
  "avg_campaign", "top_job")]

print(cluster_profile)
```

## Subscription Rate by Cluster

```{r cluster-sub-rates}
cluster_palette <- c("#2777AA", "#E03030", "#16A34A", "#D97706",
                     "#6D28D9")

ggplot(cluster_profile,
       aes(x = cluster, y = sub_rate_pct, fill = cluster)) +
  geom_col(width = 0.65, show.legend = FALSE) +
  geom_text(aes(label = paste0(sub_rate_pct, "%")),
            vjust = -0.4, size = 4, fontface = "bold") +
  scale_fill_manual(values = cluster_palette) +
  geom_hline(yintercept = mean(bank$y == "yes") * 100,
             linetype = "dashed", color = "grey40") +
  annotate("text", x = 5.4, y = mean(bank$y == "yes") * 100 + 0.5,
           label = "Overall\n11.7%", size = 3, color = "grey40") +
  labs(title    = "Subscription Rate by Cluster (k = 5)",
       subtitle = "Dashed line = overall subscription rate (11.7%)",
       x        = "Cluster",
       y        = "Subscription Rate (%)") +
  theme_minimal()
```

## Distribution of Key Variables by Cluster

```{r cluster-boxplots}
cluster_palette <- c("#2777AA", "#E03030", "#16A34A", "#D97706",
                     "#6D28D9")

# Age
ggplot(bank, aes(x = cluster, y = age, fill = cluster)) +
  geom_boxplot(outlier.alpha = 0.15, linewidth = 0.4) +
  scale_fill_manual(values = cluster_palette) +
  labs(title = "Age Distribution by Cluster",
       x = "Cluster", y = "Age") +
  theme_minimal() + theme(legend.position = "none")

# Balance
ggplot(bank, aes(x = cluster, y = balance, fill = cluster)) +
  geom_boxplot(outlier.alpha = 0.15, linewidth = 0.4) +
  scale_fill_manual(values = cluster_palette) +
  labs(title = "Account Balance by Cluster",
       x = "Cluster", y = "Balance (EUR)") +
  theme_minimal() + theme(legend.position = "none") +
  ylim(-1000, 15000)

# Duration
ggplot(bank, aes(x = cluster, y = duration, fill = cluster)) +
  geom_boxplot(outlier.alpha = 0.15, linewidth = 0.4) +
  scale_fill_manual(values = cluster_palette) +
  labs(title = "Call Duration by Cluster",
       x = "Cluster", y = "Duration (seconds)") +
  theme_minimal() + theme(legend.position = "none")
```

## Cluster Centre Heatmap

```{r cluster-heatmap}
library(reshape2)

centers_df         <- as.data.frame(km$centers)
centers_df$cluster <- paste0("Cluster ", 1:nrow(centers_df))

centers_long <- melt(centers_df,
                     id.vars       = "cluster",
                     variable.name = "Variable",
                     value.name    = "Scaled_Mean")

ggplot(centers_long,
       aes(x = Variable, y = cluster, fill = Scaled_Mean)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(Scaled_Mean, 2)), size = 3.5) +
  scale_fill_gradient2(
    low      = "steelblue",
    mid      = "white",
    high     = "tomato",
    midpoint = 0
  ) +
  labs(title    = "Cluster Centres Heatmap (k = 5)",
       subtitle = "Red = above average, Blue = below average  (all values scaled)",
       x        = "Variable",
       y        = "Cluster",
       fill     = "Scaled Mean") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

> The heatmap is the clearest summary of what makes each cluster distinct.
> Read each row as a profile: a cluster with red `balance` and red `duration`
> is a high-balance, long-call segment. A cluster with blue on all variables
> is a low-engagement, low-resource segment.

## Customer and Campaign Behaviour Segments

```{r cluster-interpretation}
cat("=== CUSTOMER SEGMENTATION SUMMARY (k = 5) ===\n\n")

for (i in 1:5) {
  cat(paste0("--- Cluster ", i, " ---\n"))
  cat("Size:              ", km$size[i], "clients  (",
      cluster_profile$pct_of_total[i], "% of total)\n")
  cat("Subscription rate: ", cluster_profile$sub_rate_pct[i], "%\n")
  cat("Avg age:           ", cluster_profile$avg_age[i], "years\n")
  cat("Avg balance:       EUR", cluster_profile$avg_balance[i], "\n")
  cat("Avg call duration: ", cluster_profile$avg_duration[i], "seconds\n")
  cat("Avg campaign:      ", cluster_profile$avg_campaign[i], "contacts\n")
  cat("Top job:           ", cluster_profile$top_job[i], "\n\n")
}
```

> Use the heatmap and profile table above to assign behavioural labels to each
cluster. The clusters represent different combinations of customer characteristics,
engagement levels, and campaign interaction patterns.
> Common segment names for this dataset:
>
>Based on the profile table generated above, clusters can be labelled according to their dominant characteristics.
## Subscription Rate Heatmap by Cluster and Job

```{r cluster-job-heatmap}
sub_by_cluster_job <- aggregate(
  (y == "yes") ~ cluster + job,
  data = bank,
  FUN  = function(x) round(mean(x) * 100, 1)
)
names(sub_by_cluster_job)[3] <- "sub_rate"

ggplot(sub_by_cluster_job,
       aes(x = cluster, y = job, fill = sub_rate)) +
  geom_tile(color = "white") +
  geom_text(aes(label = paste0(sub_rate, "%")), size = 2.8) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(title = "Subscription Rate by Cluster and Job Type",
       x     = "Cluster",
       y     = "Job Type",
       fill  = "Sub Rate (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 9))
```

> This heatmap reveals which cluster-job combinations are highest priority for
> targeting. Dark blue cells = high subscription rate for that combination.


```{r}


# ── Cluster labels (adjust numbers to match your actual cluster profiles) ──
cluster_labels <- c(
  "1" = "Young mass-market",
  "2" = "Over-contacted",
  "3" = "High-balance",
  "4" = "Engaged on call",
  "5" = "Older passive"
)

cluster_colours <- c(
  "Young mass-market" = "#E69500",
  "Older passive"     = "#2ECC71",
  "Engaged on call"   = "#3498DB",
  "High-balance"      = "#9B59B6",
  "Over-contacted"    = "#E74C3C"
)

# Add cluster name column to bank
bank$cluster_name <- factor(
  cluster_labels[as.character(bank$cluster)],
  levels = cluster_labels
)

# ── Variables: x-axis vars (columns) and y-axis vars (rows) ──────────────────
x_vars <- c("age", "balance", "campaign", "duration")
y_vars <- c("balance", "campaign", "duration", "previous")

# ── Sample for performance (optional) ────────────────────────────────────────
set.seed(123)
plot_data <- bank[sample(nrow(bank), 5000), ]

# ── Build each panel ─────────────────────────────────────────────────────────
make_panel <- function(xvar, yvar, data,
                       show_x_axis, show_y_axis,
                       show_x_label, show_y_label) {

  # Blank panel for diagonal and upper triangle
  x_idx <- which(x_vars == xvar)
  y_idx <- which(y_vars == yvar)

  # Upper triangle or diagonal: return blank
  # diagonal: xvar == yvar
  # upper: x_idx > y_idx (since y_vars is one step ahead of x_vars)
  is_blank <- (xvar == yvar) || (x_idx > y_idx)

  if (is_blank) {
    p <- ggplot() +
      theme_void() +
      theme(panel.border = element_rect(colour = "black", fill = "white",
                                        linewidth = 1))
    return(p)
  }

  p <- ggplot(data, aes(x = .data[[xvar]], y = .data[[yvar]],
                         colour = cluster_name)) +
    geom_point(size = 0.7, alpha = 0.5) +
    scale_colour_manual(values = cluster_colours, guide = "none") +
    theme_minimal(base_size = 10) +
    theme(
      panel.border     = element_rect(colour = "black", fill = NA,
                                      linewidth = 1),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(colour = "grey90"),
      plot.margin      = margin(2, 2, 2, 2)
    )

  # X axis
  if (!show_x_axis) {
    p <- p + theme(axis.text.x  = element_blank(),
                   axis.ticks.x = element_blank())
  }
  if (!show_x_label) {
    p <- p + labs(x = NULL)
  }

  # Y axis
  if (!show_y_axis) {
    p <- p + theme(axis.text.y  = element_blank(),
                   axis.ticks.y = element_blank())
  }
  if (!show_y_label) {
    p <- p + labs(y = NULL)
  }

  p
}

# ── Assemble the 4x4 grid ────────────────────────────────────────────────────
n_x <- length(x_vars)   # 4 columns
n_y <- length(y_vars)   # 4 rows

plot_list <- list()

for (row_i in seq_along(y_vars)) {
  for (col_j in seq_along(x_vars)) {

    yvar <- y_vars[row_i]
    xvar <- x_vars[col_j]

    show_x <- (row_i == n_y)          # only bottom row shows x axis text
    show_y <- (col_j == 1)            # only left column shows y axis text

    p <- make_panel(
      xvar        = xvar,
      yvar        = yvar,
      data        = plot_data,
      show_x_axis = show_x,
      show_y_axis = show_y,
      show_x_label = FALSE,
      show_y_label = FALSE
    )

    plot_list[[length(plot_list) + 1]] <- p
  }
}

# ── Column headers (top) ─────────────────────────────────────────────────────
col_headers <- lapply(x_vars, function(v) {
  ggplot() +
    annotate("text", x = 0.5, y = 0.5, label = v,
             size = 5, fontface = "bold") +
    theme_void() +
    theme(
      panel.border = element_rect(colour = "black", fill = "white",
                                  linewidth = 1)
    )
})

# ── Row labels (right side) ──────────────────────────────────────────────────
row_labels <- lapply(y_vars, function(v) {
  ggplot() +
    annotate("text", x = 0.5, y = 0.5, label = v,
             size = 5, fontface = "bold", angle = 270) +
    theme_void() +
    theme(
      panel.border = element_rect(colour = "black", fill = "white",
                                  linewidth = 1)
    )
})

# ── Legend ───────────────────────────────────────────────────────────────────
legend_data <- data.frame(
  cluster_name = factor(names(cluster_colours), levels = names(cluster_colours)),
  x = 1, y = 1
)

legend_plot <- ggplot(legend_data, aes(x, y, colour = cluster_name)) +
  geom_point(size = 3) +
  scale_colour_manual(values = cluster_colours, name = NULL) +
  theme_void() +
  theme(
    legend.position  = "bottom",
    legend.direction = "horizontal",
    legend.text      = element_text(size = 11)
  )

legend_grob <- cowplot::get_legend(legend_plot)

# ── Final assembly with patchwork ────────────────────────────────────────────

# Top row: blank corner + 4 column headers + blank corner
top_row <- (plot_spacer() | wrap_elements(col_headers[[1]]) |
              wrap_elements(col_headers[[2]]) |
              wrap_elements(col_headers[[3]]) |
              wrap_elements(col_headers[[4]]) |
              plot_spacer()) +
  plot_layout(widths = c(0.1, 1, 1, 1, 1, 0.15))

# Data rows: 4 scatter panels + right label per row
make_row <- function(row_i) {
  panels <- plot_list[((row_i - 1) * n_x + 1):(row_i * n_x)]
  row_label <- wrap_elements(row_labels[[row_i]])
  (wrap_elements(panels[[1]]) | wrap_elements(panels[[2]]) |
     wrap_elements(panels[[3]]) | wrap_elements(panels[[4]]) |
     row_label) +
    plot_layout(widths = c(1, 1, 1, 1, 0.15))
}

final_plot <- top_row /
  make_row(1) /
  make_row(2) /
  make_row(3) /
  make_row(4) /
  wrap_elements(legend_grob) +
  plot_layout(heights = c(0.30, 1, 1, 1, 1, 0.18))

final_plot
```

## Cluster Silhouette Plot

```{r cluster-silhouette-plot}
cluster_palette <- c("#2777AA", "#E03030", "#16A34A", "#D97706",
                     "#6D28D9")

sil <- silhouette(km$cluster[sample_idx], dist(cluster_sample))

fviz_silhouette(sil,
                palette = cluster_palette,
                ggtheme = theme_minimal()) +
  labs(title    = "Silhouette Plot — Cluster Quality (k = 5)",
       subtitle = "Based on sample of 5,000 clients. Positive = well-assigned.")
```

> Clusters where most bars are positive and tall are internally cohesive and
> well-separated from neighbours. Negative bars indicate observations the algorithm
> assigned to the wrong cluster — a small number is acceptable.

## Business Implications of Segmentation

```{r segment-summary}
cat("=== BUSINESS IMPLICATIONS (k = 5) ===\n\n")

cat("Cluster with HIGHEST subscription rate:\n")
best_cluster  <- cluster_profile$cluster[which.max(cluster_profile$sub_rate_pct)]
cat("  Cluster", as.character(best_cluster),
    "— Sub rate:", max(cluster_profile$sub_rate_pct), "%\n")
cat("  PRIORITY TARGET — concentrate campaign effort here.\n\n")

cat("Cluster with LOWEST subscription rate:\n")
worst_cluster <- cluster_profile$cluster[which.min(cluster_profile$sub_rate_pct)]
cat("  Cluster", as.character(worst_cluster),
    "— Sub rate:", min(cluster_profile$sub_rate_pct), "%\n")
cat("  Lowest return on telemarketing investment.\n\n")

cat("Ranked subscription rates by cluster:\n")
ranked <- cluster_profile[order(cluster_profile$sub_rate_pct,
                                decreasing = TRUE),
                          c("cluster", "sub_rate_pct", "n",
                            "avg_age", "avg_balance", "avg_duration")]
print(ranked, row.names = FALSE)

```

> **Key business insight:** The 5 clusters have substantially different subscription
> rates. The highest-rate cluster represents the ideal target profile — clients who
> share characteristics with previous subscribers. The lowest-rate cluster represents
> wasted calls. Directing campaign effort based on cluster membership, rather than
> calling randomly, has the potential to improve subscription yield significantly.

> K-Means reveals 5 distinct customer and campaign behaviour segments with
> meaningfully different subscription rates. The results suggest that engagement
> patterns, particularly call duration and campaign intensity, are associated with
> different levels of subscription success.
---


*End of Report*
